---
title: "Unmasking the Numbers: An Analysis of the BNT162b2 mRNA Covid-19 Vaccine Efficacy Rate"
subtitle: "Frequentist and Bayesian Analysis of Polack et. al."
graphics: yes
author: "Nathan Dennis, Aarav Dewangan, Maxwell Wang"
date: "2023-05-26"
output: 
        pdf_document:
         toc: false
         number_sections: true
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{multicol,graphicx,hyperref,xcolor}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

\begin{table}[ht]
\centering
\caption{Sample Table}
\begin{tabular}{|c|c|c|}
\hline
Groups & Cases & No. of Subjects \\
\hline
BNT162b2 & 8 & 17,411 \\
\hline
Placebo & 162 & 17,511 \\
\hline
Total & 170 & 34,922 \\

\hline
\end{tabular}
\end{table}


## Abstract


## Keywords


## Introducton

## Something else

We will define null hypothesis $H_0$: $\pi_p = \pi_v$, which represents the vaccine having no affect on getting infected with Covid-19 compared with the placebo.

We will define the one-sided alternative hypothesis $H_a$: $\pi_v < \pi_p$ representing that vaccinated patients are less likely to get infected with Covid-19 than patients who received the placebo.

## Statistical Methods

We will use parameters $\pi_p$ to represent the likelihood of being infected with Covid-19 after taking the placebo, and $\pi_v$ to represent the likelihood of getting Covid-19 after receiving the vaccine.

We will define null hypothesis $H_0$: $\pi_p = \pi_v$, which represents the vaccine having no affect on getting infected with Covid-19 compared with the placebo.

We will define the one-sided alternative hypothesis $H_a$: $\pi_v < \pi_p$ representing that vaccinated patients are less likely to get infected with Covid-19 than patients who received the placebo.

Let X denote the total number of covid cases among $n_1$(17411) subjects randomly assigned to the BNT162b2 group and let Y denote the total number of covid cases among $n_2$(17511) subjects randomly assigned to the placebo group.\

$X \sim Binom(n_1,\pi_v)$\
$Y \sim Binom(n_2,\pi_p)$\
The parameter of interest is the BNT162b2 efficacy:\
$\psi=\frac{\pi_p- \pi_v}{\pi_p}$\

Let $\pi_v=\mbox{probability of covid 19 in vaccine}$\
Let $\pi_p=\mbox{probability of covid 19 in placebo}$\
From the data table, we see that:\
$\hat \pi_v=\frac{8}{17411}$\
$\hat \pi_p=\frac{162}{17511}$\
$\hat\psi=\frac{\hat \pi_p-\hat \pi_v}{\hat \pi_p}$\
$\hat\psi=\frac{\frac{162}{17511}-\frac{8}{17411}}{\frac{162}{17511}}$\
$\hat\psi=\frac{446749}{470097}$\
$\hat\psi=0.950334$\

We see that vaccine reduces the size of the Covid 19 by 95\% compared to placebo.\


```{r label ="bayesian", include = FALSE, cache = TRUE}


```


## Results


## Discussion/Conclusion


## References


## Appendix

```{r ref.label="bayesian", echo=TRUE, eval=F}
```
