---
title: "Unmasking the Numbers: An Analysis of the BNT162b2 mRNA Covid-19 Vaccine Efficacy Rate"
subtitle: "Frequentist and Bayesian Analysis of Polack et. al."
graphics: yes
author: "Nathan Dennis, Aarav Dewangan, Maxwell Wang"
date: "2023-05-26"
output: 
        pdf_document:
         toc: false
         number_sections: false
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{multicol,graphicx,hyperref,xcolor}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=F}
library(ggplot2)
library(dplyr)
library(gridExtra)
```


\begin{table}[ht]
\centering
\caption{Sample Table}
\begin{tabular}{|c|c|c|}
\hline
Groups & Cases & No. of Subjects \\
\hline
BNT162b2 & 8 & 17,411 \\
\hline
Placebo & 162 & 17,511 \\
\hline
Total & 170 & 34,922 \\

\hline
\end{tabular}
\end{table}


## Abstract

# Background


## Keywords

COVID-19, Vaccine, Efficacy, Immunization, Analysis

## Introducton

<include> BNT162b2 mRNA Covid-19 Vaccine, which we will refer to subsequently as simply "vaccine"

\pagebreak
## Large Sample T-Test

We'll assume the infection status of each patient is a Bernoulli Random Variable, where the success rate $\pi$ determines the probability that a patient contracted Covid-19 during the study.

Let us define $X_1, X_2, \dots, X_n \overset{i.i.d.}{\sim} \mbox{Bernoulli}(\pi_p)$ and $Y_1, Y_2, \dots, Y_m \overset{i.i.d.}{\sim} \mbox{Bernoulli}(\pi_v)$. $X_i$'s represent the infection status of patients who received the placebo, and $Y_i$'s represent the infection status of patients who received the COVID-19 vaccine.

<TODO>?
The efficacy rate is defined as $1 - \frac{\pi_v}{\pi_p}$, so let's define the random variable Z to represent the estimated efficacy rate from our data, where $Z = 1 - \frac{\bar{Y}}{\bar{X}}$. We want to see whether or not the efficacy rate exceeds the 30\% as required by the FDA.

From our data, our estimated efficacy
$\hat\psi=\frac{\hat \pi_p-\hat \pi_v}{\hat \pi_p}$
$\hat\psi=0.950334$
<TODO>?

We'll assume that the sample size is sufficiently large, since $n = 17511$ and $m = 17411$. Then the Central Limit Theorem should hold, so then $\bar{X} \sim \mbox{Norm}(\mu_x, \frac{\sigma_x}{\sqrt{n}})$ and $\bar{Y} \sim \mbox{Norm}(\mu_y, \frac{\sigma_y}{\sqrt{m}})$. The mean of a Bernoulli random variable with probability $\pi$ is $\pi$ and the variance is $\pi(1 - \pi)$.

Therefore, $\bar{X} \sim \mbox{Norm}(\pi_p, \sqrt{\frac{\pi_p(1 - \pi_p)}{n}})$ and $\bar{Y} \sim \mbox{Norm}(\pi_v, \sqrt{\frac{\pi_v(1 - \pi_v)}{m}})$.

$xbar - ybar +- z_alpha/2 sqrt(sigma1^2/n + sigma2^2/m)$

$xbar Sim Norm (mu1, sigma1 / sqrt(n))$

We will define null hypothesis $H_0$: $Z = .3$, which represents the vaccine having 30\% efficacy.
We will define the one-sided alternative hypothesis $H_a$: $Z > .3$ representing that the vaccine has an efficacy greater than 30\%.

We will define null hypothesis $H_0$: $\pi_p = \pi_v$, which represents the vaccine having no affect on getting infected with Covid-19 compared with the placebo.

We will define the one-sided alternative hypothesis $H_a$: $\pi_v < \pi_p$ representing that vaccinated patients are less likely to get infected with Covid-19 than patients who received the placebo.


\pagebreak
=======

## Statistical Methods
## Bayesian Inference

### Definitions

We will use parameters $\pi_p$ to represent the likelihood of being infected with Covid-19 after taking the placebo, and $\pi_v$ to represent the likelihood of getting Covid-19 after receiving the vaccine.

We will define null hypothesis $H_0$: $\pi_p = \pi_v$, which represents the vaccine having no affect on getting infected with Covid-19 compared with the placebo.

We will define the one-sided alternative hypothesis $H_a$: $\pi_v < \pi_p$ representing that vaccinated patients are less likely to get infected with Covid-19 than patients who received the placebo.

Let X denote the total number of laboratory confirmed Covid-19 cases among $n_1$(17411) subjects randomly assigned to the BNT162b2 group and let Y denote the total number of laboratory confirmed Covid-19 cases among $n_2$(17511) subjects randomly assigned to the placebo group.\

$X \sim Binom(n_1,\pi_v)$\
$Y \sim Binom(n_2,\pi_p)$\
The parameter of interest is the BNT162b2 efficacy:\
$\psi=\frac{\pi_p- \pi_v}{\pi_p}$\

Let $\pi_v=\mbox{probability of laboratory confirmed Covid-19 cases in vaccine}$\
Let $\pi_p=\mbox{probability of laboratory confirmed Covid-19 cases in placebo}$\
From the data table, we see that:\
$\hat \pi_v=\frac{8}{17411}$\
$\hat \pi_p=\frac{162}{17511}$\
$\hat\psi=\frac{\hat \pi_p-\hat \pi_v}{\hat \pi_p}$\
$\hat\psi=\frac{\frac{162}{17511}-\frac{8}{17411}}{\frac{162}{17511}}$\
$\hat\psi=\frac{446749}{470097}$\
$\hat\psi=0.950334$\

We see that vaccine reduces the rate of Covid 19 infection by 95\% compared to the placebo, which matches that of Polack et. al.\

Specifically, assuming the trial will run until X+Y=s Covid cases observed, we consider the distribution of :\
$T:\mbox{number of covid cases in BNT162b2 group from these s cases}$\
The distribution of T well approximated by binomial for the BNT162b2 group study. Therefore:\
$T \sim Binom(s,\pi)$, where $\pi=P(\mbox{BNT162b2 group}|\mbox{laboratory confirmed Covid-19 cases})$\


$\pi=P(\mbox{BNT162b2 group}|\mbox{laboratory confirmed Covid-19 cases})$\
 Using Bayes Theorem to simplify:\
$\pi=\frac{P(\mbox{laboratory confirmed Covid-19 cases}|\mbox{BNT162b2 group})P(\mbox{BNT162b2 group})}{P(\mbox{laboratory confirmed Covid-19 cases})}$\

Finding $P(\mbox{laboratory confirmed Covid-19 cases})$:\
$P(\mbox{laboratory confirmed Covid-19 cases})=\pi_v$\

Finding $P(\mbox{BNT162b2 group})$:\
$P(\mbox{BNT162b2 group})=\frac{n_1}{n_1+n_2}$\

Finding $P(\mbox{laboratory confirmed Covid-19 cases})$:\
$P(\mbox{laboratory confirmed Covid-19 cases})$\
$=\frac{\mbox{Total no. of confirmed Covid-19 cases in vaccine + Total no. of confirmed Covid-19 cases in placebo} }{\mbox{Total number of subjects in vaccine + Total number of subjects in plcebo }}$\
$=\frac{n_1\pi_v+n_2\pi_p}{n_1+n_2}$\
$P(\mbox{laboratory confirmed Covid-19 cases})=\frac{n_1\pi_v+n_2\pi_p}{n_1+n_2}$\

Therefore:\
$\pi=\frac{\pi_v\cdot \frac{n_1}{n_1+n_2}}{\frac{n_1\pi_v+n_2\pi_p}{n_1+n_2}}$\
$=\frac{\frac{n_1\pi_v}{n_1+n_2}}{\frac{n_1\pi_v+n_2\pi_p}{n_1+n_2}}$\
$=\frac{n_1\pi_v}{n_1+n_2}$\
$\pi=\frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p}$\
When $n_1\approx n_2$, we say randomization is 1:1 and are able to simplify $\pi$ further:\
$\pi=\frac{\pi_v}{\pi_v+\pi_p}$\

Thus $T \sim Binom(s,\pi=\frac{\pi_v}{\pi_v+\pi_p})$\
Our parameter of interest is:\
$\psi=\frac{\pi_p- \pi_v}{\pi_p}=1-\frac{\pi_v}{\pi_p}$\
We know $\pi=\frac{\pi_v}{\pi_v+\pi_p}$\
$\implies \pi(\pi_v+\pi_p)=\pi_v$\
$\implies \pi\pi_v+\pi\pi_p=\pi_v$\
$\implies \pi\pi_p=\pi_v(1-\pi)$\
$\implies \pi_v=\frac{\pi\pi_p}{1-\pi}$\

$\psi=1-\frac{\pi_v}{\pi_p}$\
$\implies 1-\frac{\pi\pi_p}{(1-\pi)\pi_p}$\
$\implies 1-\frac{\pi}{1-\pi}$\
$\implies \frac{1-2\pi}{1-\pi}$\
$\psi=\frac{1-2\pi}{1-\pi}$\

### Prior from Polack et. al.

We define $T \sim Binom(170,\pi)$ where $\pi=\frac{\pi_v}{\pi_v+\pi_p}$. We will also use the prior distribution for $\pi$, $g(\pi) = Beta(0.700102,1)$ as Polack et. al. did. In the article, it was also mentioned that the 95% confidence interval for $\pi$ is (0.005, 0.964) and the corresponding 95\% interval for $\psi$ is (-26.2, 0.995).\

We know beta is a conjugate prior in the binomial model. In the other words, posterior distribution of $\pi$ is also a beta distribution.\

$h(\pi_0|t)=Beta(0.700102+8,170-8+1)$\
$h(\pi_0|t)=Beta(8.700102,163)$\

We can find the 95% Bayesian credible interval for $\pi_0$, where we transform this for the vaccine efficacy, $\psi$. We calculate the 95% credible interval for $\pi_0$, [0.023194, 0.08799]. We can convert this to create an interval for $\psi$ using the fact that $\psi = \frac{1-2\pi}{1-\pi}$. For the lower bound,

\begin{align*}
\frac{1-2 \cdot 0.023194}{1-0.023194} &= 0.903520
\end{align*}

For the upper bound,

\begin{align*}
\frac{1-2 \cdot 0.08799}{1-0.08799} &= 0.976255
\end{align*}


```{r label ="bayesian", include = FALSE, cache = TRUE}

pi_ci <- qbeta(p=c(0.025,0.975), 8.700102, 163)


upper_ci_prior1 <- (1-2*pi_ci[1])/(1-pi_ci[1])
lower_ci_prior1 <- (1-2*pi_ci[2])/(1-pi_ci[2])
```


We obtained a 95% Bayesian credible interval reported as [90.3%, 97.6%]. This interval does not contain the value 30% and is much larger than this, so we conclude the vaccine efficacy is larger than 30%.



### Alternative Beta Prior, Further Pessimistic Approach

Polack et. al. cites a prior distribution for $\pi$ of $\mbox{Beta}(0.700102, 1)$ which is intended to be centered at $\pi = 0.4118$ corresponding to a $\psi$ of 30\%. The authors cite this as a pessimistic estimate. However, we will try a more pessimistic prior, using a 0\% vaccine efficacy rate. A "pessimistic" prior would represent a more skeptical approach in regards to vaccine efficacy, where we assume a vaccine efficacy of 0\%. In many cases it is argued that pessimistic priors are rarely necessary since researchers are motivated to conduct a trial with a belief that the vaccine will be beneficial, however this is a flawed argument since in many cases prior beliefs about some intervention could prove to be inaccurate, causing significantly more harm. (Zampieri, Fernando G, et al.) Due to this, using a prior implying the vaccine efficacy rate is 0\% is a safe option to eliminate the possiblity of substantial harm.

When $\psi = 0$, $\pi = \frac{\psi - 1}{\psi - 2} = \frac{1}{2}$. This corresponds to the $\mbox{Beta}(1, 1)$ prior, which is uniform. The prior in this case can actually be thought of as a uniform distribution, Unif(0,1). Under the prior, the 95\% CI for $\pi$ is $[0.025, 0.975]$ which corresponds to a 95\% CI for $\psi$ being $[-38, 0.974359]$. This implies that we have enough room for uncertainty again,
```{r prior 1 uh}
pi_ci_3 <- qbeta(p=c(0.025,0.975), 1, 1)
pi_ci_3
```

Well use the Beta conjugate, to find our posterior distribution, where we define \
$h(\pi_0 | t) = \mbox{Beta}(1 + 8, 170 - 8 + 1) = \mbox{Beta}(9, 163)$.

Using R software we calculate the 95% credible interval for $\pi_0$ to be [0.024346, 0.090099]. We again convert this to create an interval for $\psi$ using the equality $\psi = \frac{1-2\pi}{1-\pi}$. So our 95\% interval for $\psi$ is (0.900979, 0.975047). This interval is above and does not contain the value of 0.3, so we conclude the vaccine efficacy rate is greater than 30\%.

```{r code prior2}
pi_ci2 <- qbeta(p=c(0.025,0.975), 9, 163)

upper_ci_prior2 <- (1-2*pi_ci2[1])/(1-pi_ci2[1])
lower_ci_prior2 <- (1-2*pi_ci2[2])/(1-pi_ci2[2])
upper_ci_prior2
lower_ci_prior2
```


### Poisson Prior?

So far we have constructed beta priors and posteriors from the Binomial distribution. Suppose now we can construct a Poisson distribution using the previous binomial distribution defined as $T \sim Binom(170, \pi)$. A good approximation for the binomial when the number of samples is large would be a poisson distribution with $\lambda = 170\pi$. We can denote $U \sim Poisson(\lambda = 170\pi)$, where $\lambda$ represents the rate of covid cases in the vaccine group over 170 positive cases and U represents the number of covid cases in the vaccine group over the 170 positive cases.

For a Poisson distribution, we utilize Jeffreys prior for the rate parameter $\lambda \geq 0$, which states that $p(\lambda) = \sqrt{\frac{1}{\lambda}}$. However, this can also be thought of as a $Gamma(\frac{1}{2},0)$ distribution, which implies the posterior takes the form $\lambda|u \sim Gamma(\frac{1}{2}+u,170)$.

```{r}
qgamma(0.025, 1/2+8, 170)
qgamma(0.975, 1/2+8, 170)

upper_ci_prior2 <- (1-2*0.02224761)/(1-0.02224761)
lower_ci_prior2 <- (1-2*0.08879709)/(1-0.08879709)
upper_ci_prior2
lower_ci_prior2
```

Proof for Prior:



\pagebreak
## Max-Likelihood

We want to make a maximum-likelihood estimator for $\psi$, $\widehat{\psi^{mle}}$, given the observed data. Let us start by writing the likelihood function. We have shown previously that $T\sim Binom(170,\pi)$, where the distribution would be defined as: $f(t|\pi) = {170\choose t} \pi^t \cdot (1 - \pi)^{170-t}$. We need to reparamaterize this with $\psi$, which represents the number of patients who contracted COVID who had initially been in the vaccinated group rather than the placebo. We use the fact that $\psi = \frac{1-2\pi}{1-\pi}$, where we can rearrange terms:
$\textbf{Lemma 1.}$
\begin{align*}
\psi &= \frac{1-2\pi}{1-\pi} \\
(1-\pi)\psi &= 1-2\pi \\
\psi - \pi \psi &= 1-2\pi \\
2\pi - \pi \psi &= 1 - \psi \\
\pi &= \frac{1 -\psi}{2-\psi}
\end{align*}
Plugging in this value for $\pi$, we now have that $f(t|\psi) = {170\choose t} (\frac{1 -\psi}{2-\psi})^t \cdot (1 - \frac{1 -\psi}{2-\psi})^{170-t}$. So the likelihood function is as follows: 
\begin{align*}
L(\psi) &= f(t | \psi) \\;
&= {170\choose t} (\frac{1 -\psi}{2-\psi})^t \cdot (1 - \frac{1 -\psi}{2-\psi})^{170-t} & \text{Lemma 1} \\
&= {170\choose t} (\frac{1 -\psi}{2-\psi})^t \cdot (\frac{1}{2-\psi})^{170-t}
\end{align*}
Next, we compute the log-likelihood function by taking the natural log of the likelihood function. We denote this log-likelihood as $\ell(\psi) = \ln(L(\psi)) = \ln(f(t | \psi))$.
\begin{align*}
\ell(\psi) &= \ln(f(t | \psi)) \\
&= \ln \left({170\choose t} (\frac{1 -\psi}{2-\psi})^t \cdot (\frac{1}{2-\psi})^{170-t} \right) \\
&= \ln{170\choose t} + \ln\left((\frac{1 -\psi}{2-\psi})^t \right) + \ln \left((\frac{1}{2-\psi})^{170-t} \right) \\
&= \ln{170\choose t} + t\ln(\frac{1 -\psi}{2-\psi}) + (170-t)\ln(\frac{1}{2-\psi}) 
\end{align*}
We have calculated the log-likelihood $\ell(\psi) = \ln{170\choose t} + t\ln(\frac{1 -\psi}{2-\psi}) + (170-t)\ln(\frac{1}{2-\psi})$. To find the maximum of the likelihood function, we now take derivative of the log-likelihood and set it to zero. First, take the derivative with respect to $\psi$.
\begin{align*}
\ell^\prime (\psi) &= \frac{d}{d\psi} \ell(\psi) \\
&= \frac{d}{d\psi} \left(\ln{170\choose t} + t\ln(\frac{1 -\psi}{2-\psi}) + (170-t)\ln(\frac{1}{2-\psi}) \right) \\
&= t \frac{2 -\psi}{1-\psi} \cdot (-\frac{1}{(2 -\psi)^2}) + (170-t)(2-\psi) \cdot \frac{1}{(2 -\psi)^2} \\
&= -\frac{t}{(1-\psi)(2 -\psi)} + (170-t) \cdot \frac{1}{(2 -\psi)}
\end{align*}
Given this derivative we can set it equal to 0. Also note that since T represents the number of covid cases in the vaccine group within the 170 cases, we have given in the dataset t=8 and can plug this value in to solve.

\begin{align*}
0 &= -\frac{t}{(1-\psi)(2 -\psi)} + (170-t) \cdot \frac{1}{(2 -\psi)}\\
\frac{8}{(1-\psi)(2 -\psi)} &= (170-8)\frac{1}{(2 -\psi)}\ \ \text{Set t = 8, given} \\
\frac{8}{162} &= \frac{(1-\psi)(2 -\psi)}{(2 -\psi)}\\
\frac{8}{162} &= 1-\psi\\
\psi &= 1 - \frac{8}{162} = \frac{77}{81}
\end{align*}

We have calculated the critical point for the maximum likelihood estimator to be $\psi = \frac{77}{81}$. To check whether our value of $\psi = \frac{77}{81}$ is a global maximum, we will take the second derivative of our log-likelihood function.
\begin{align*}
\ell^{\prime \prime}(\psi) &= \frac{d}{d\psi} \ell^\prime(\psi)\\
&= \frac{d}{d\psi} \left[\frac{-t}{(1-\psi)(2 - \psi)} + \frac{(170-t)}{(2-\psi)} \right]\\
&= \frac{d}{d\psi} [-t (2 - 3\psi + \psi^2)^{-1} + (170 - t)(2-\psi)^{-1}] \\
&= -t\cdot-1\cdot(2-3\psi+\psi^2)^{-2}(-3 + 2\psi) + (170 - t) \cdot -1 \cdot (2-\psi)^{-2} \cdot -1 \\
&= \frac{t(-3 + 2\psi)}{(2-3\psi+\psi^2)^2} + \frac{(170 - t)}{(2-\psi)^2}\\
\end{align*}

```{r label="mle 2nd derivative"}
t <- 8
psi <- 77/81
second_deriv <- ((t * (-3 + 2 * psi))/((2 - 3 * psi + psi ** 2) ** 2)) + ((170 - t) / ((2 - psi) ** 2))
```

As mentioned previously, from our data, $t = 8$ because $8$ out of $170$ total COVID cases came from the vaccinated group. So, the second derivative at $\psi = 77/81$ is $\frac{t(-3 + 2\psi)}{(2-3\psi+\psi^2)^2} + \frac{(170 - t)}{(2-\psi)^2}$ =
`r second_deriv`

Since the second derivative is negative, that means we have a local maximum at $\psi = 77/81$. Because we only found one point where the first derivative is 0, we only need to check the likelihood at the other two critical points, which are the boundaries. Specifically, that would be $\psi = 0$ and $\psi = 1$. 

```{r label="mle critical points"}
t = 8
likelihood_fn <- function(p) {
  log(choose(170, t)) + t * log((1 - p)/(2 - p)) + (170 - t) * log(1/(2 - p))
}
likelihood_fn(77/81)
likelihood_fn(0)
likelihood_fn(1)
```

Since the likelihood function is the greatest at $\psi_0 = \frac{77}{81}$, this confirms that $\widehat{\psi_0^{mle}} = \frac{77}{81}$ is indeed the global maximizer for the likelihood function. Thus it is our maximum likelihood estimator for vaccine efficacy.

We can also construct a Wald confidence interval for $\psi_0$ at $\alpha = 0.05$. The formula for this confidence interval is $\widehat{\psi_0^{mle}} \pm z_{\alpha / 2} \sqrt{\frac{1}{-\ell^{\prime \prime} (\widehat{\psi_0}^{mle})}}$. We have calculated all the values we need,

\begin{align*}
\widehat{\psi_0^{mle}} \pm z_{\alpha / 2} \sqrt{\frac{1}{-\ell^{\prime \prime} (\widehat{\psi_0}^{mle})}} &= \frac{77}{81} \pm z_{0.05 / 2} \sqrt{\frac{1}{3126.123529}}
\end{align*}

```{r label="wald"}
t <- 8
psi <- 77/81
second_deriv <- ((t * (-3 + 2 * psi))/((2 - 3 * psi + psi ** 2) ** 2)) + ((170 - t) / ((2 - psi) ** 2))
lo <- psi + qnorm(0.05/2) * sqrt(1 / (-second_deriv))
hi <- psi - qnorm(0.05/2) * sqrt(1 / (-second_deriv))
lo
hi
```

We conclude that $\widehat{\psi_0^{mle}} = \frac{77}{81}$. Furthermore, our 95\% Wald confidence interval for $\psi_0$ is [`r lo`, `r hi`]. In other words, we are 95\% confident that the true vaccine efficacy rate lies between `r lo` and `r hi`. Since the entire confidence interval is well above the FDA's requirement of 30\%, we conclude that the vaccine is sufficiently effective.


\pagebreak
## Bootstrap Confidence Interval

We'll assume the infection status of each patient is a Bernoulli Random Variable, where the success rate $\pi$ determines the probability that a patient contracted Covid-19 during the study.

Let us define $X_1, X_2, \dots, X_n \overset{i.i.d.}{\sim} \mbox{Bernoulli}(\pi_p)$ and $Y_1, Y_2, \dots, Y_m \overset{i.i.d.}{\sim} \mbox{Bernoulli}(\pi_v)$. $X_i$'s represent the infection status of patients who received the placebo, and $Y_i$'s represent the infection status of patients who received the COVID-19 vaccine. Then $\sum_{i = 1}^n X_i \sim \mbox{Binom}(\pi_v)$ and $\sum_{i = 1}^m X_i \sim \mbox{Binom}(\pi_p)$ and we can approximate $\pi_p = \bar{X}$ and $\pi_v = \bar{Y}$, which we can find by dividing the sum (a binomial random variable) by $n$ or $m$, respectively.

Using $\hat{\pi_p}$ and $\hat{\pi_v}$ from the data as the parameters for our binomial distributions, we will run a parametric bootstrap simulation with B = 10,000 trials and construct a corresponding bootstrap confidence interval. Recall vaccine efficacy $\psi$ is defined as $\frac{\pi_p - \pi_v}{\pi_p} = 1 - \frac{\pi_v}{\pi_p}$. So in terms of our bootstrap, $\psi \approx 1 - \frac{\bar{X}}{\bar{Y}}$.

```{r label ="bootstrap", include = FALSE, cache = TRUE}
# Requirements
set.seed(342)

# Data from study
n <- 17411
m <- 17511
pi_v_hat <- 8/n
pi_p_hat <- 162/m

# Helper function to compute psi from bootstrapped data
get_efficacy <- function(x_bar, y_bar) {
  1 - (x_bar / y_bar)
}

# Number of trials, B
B <- 10000
# Vaccine efficacy psi for our dataset
psi <- get_efficacy(pi_v_hat, pi_p_hat)

# Run bootstrap
boot_sim <- lapply(1:B, FUN = function(i){ 
    # generate a re-sample from known distribution
    x_bar = rbinom(1, n, pi_v_hat) / n
    y_bar = rbinom(1, m, pi_p_hat) / m
    
    #calculate bootstrapped efficacy and return value as dataframe
    psi_bootstrapped = get_efficacy(x_bar, y_bar)
    data.frame(psi_bootstrapped)
   })

boot_sim_psi <- do.call(rbind, boot_sim)

# Large histogram
p1 <- ggplot(data=boot_sim_psi, 
       mapping = aes(psi_bootstrapped))+
  geom_histogram(binwidth =0.01)+
  labs(title="Histogram of 10,000 Bootstrapped Vaccine Efficacy Rates",x="VE",y="Count")+
  xlim(0, 1)

# Detailed histogram with normality
p2_1 <- ggplot(data=boot_sim_psi, 
       mapping = aes(psi_bootstrapped))+
  geom_histogram(binwidth =0.005)+
  labs(title="Histogram of 10,000 Bootstrapped Vaccine Efficacy Rates",x="VE",y="Count")+
  xlim(0.8, 1)

# Normal qq
p2_2 <- ggplot(data = boot_sim_psi,
             mapping = aes(sample = psi_bootstrapped)) +
  stat_qq(distribution = qnorm) + 
  stat_qq_line(distribution = qnorm) + 
  labs(title = "Normal probability plot",
       subtitle = "of bootstrapped distribution")

# Center of bootstrapped confidence interval for psi
bootstrapped_center <-mean(boot_sim_psi$psi_bootstrapped)

# Compute Bootstrap confidence interval
# Percentile method
percentile_method<-boot_sim_psi %>% summarize(lower = quantile(psi_bootstrapped,.025),
                      upper = quantile(psi_bootstrapped, .975))
percentile_method # See output

median <- quantile(boot_sim_psi$psi_bootstrapped, 0.5)

# Bias adjusted Standard bootstrap method
standard_bootstrap<-boot_sim_psi %>% summarise(boot_mean=mean(psi_bootstrapped),
                          boot_se = sd(psi_bootstrapped),
                          bias = boot_mean-psi,
                          lower = psi + qnorm(p=0.025) * boot_se,
                          upper = psi - qnorm(p=0.025) * boot_se)

standard_bootstrap # See output

summary_table <- matrix(c(percentile_method$lower,percentile_method$upper,standard_bootstrap$lower,standard_bootstrap$upper),ncol=2,byrow=TRUE)
colnames(summary_table) <- c("Lower","Upper")
rownames(summary_table) <- c("Percentile Method","Standard Bootstrap")
summary_table <- as.table(summary_table)
summary_table
```

```{r figure 3.1}
grid.arrange(p2_1, p2_2)
```

Based on the Normal Q-Q plot, we conclude that the bootstrapped distribution is not nearly normal as the tail ends of the Q-Q plot differ from the expected values should our bootstrapped values of $\psi$ come from a normal distribution. Therefore, we should prefer the percentile method 95\% confidence interval for $\psi$ instead. In other words, we are 95\% confident that the true vaccine efficacy is between [0.9110567, 0.9817138], with our median bootstrapped $\psi = 0.95145$. Our confidence interval lies above 30\% so there is sufficient statistical evidence to conclude that the vaccine is sufficiently effective <TODO Phrasing>



\pagebreak
## Results


## Discussion/Conclusion


## References

Zampieri, Fernando G, et al. “Using Bayesian Methods to Augment the Interpretation of Critical Care Trials. an Overview of Theory and Example Reanalysis of the Alveolar Recruitment for Acute Respiratory Distress Syndrome Trial.” American Journal of Respiratory and Critical Care Medicine, 1 Mar. 2021, www.ncbi.nlm.nih.gov/pmc/articles/PMC7924582/. 

https://myweb.uiowa.edu/pbreheny/uk/teaching/701/notes/1-22.pdf

https://people.bath.ac.uk/masss/ma40189/homeworks/MA40189-solution5.html#:~:text=2%20%3D%20nθ.-,The%20Jeffreys%20prior%20is%20then%20f(θ)%20∝%20√n,nˉx%2Cn).


\pagebreak
## Appendix

```{r ref.label="bayesian", echo=TRUE, eval=F}
```


## Challenge
```{r}
posterior_median_pi <- qbeta(0.5, 8.700102, 163)
posterior_median_pi

(1-0.04893234*2)/(1-0.04893234)

```

$P \left(\frac{1-\psi}{2-\psi} < 0.04893234|t \right) = 0.5$ which simplifies too $P(\psi>0.9485501|t)=0.5$

Posterior median of $\psi$ is 0.9485501. We want to test the null hypothesis for $H_0:\psi_0 = 30\%$ and the alternative  $H_1:\psi_0 > 30\%$. We calculate the posterior probability that $P(\psi_0 \leq 0.3|t)$